### 2017-04-04 paper/sexSelectionSpacing-ver1.tex
Shorten the theory section to focus on spacing only

% sex-selective Abortion
% Begun originally: 01/11/02
% Begun.: 2016-05-12
% Edited: 2016-11-07

@@

% \setcounter{page}{-1}
% \maketitle
% \thispagestyle{empty}
% 
% % \setcounter{page}{0}
% \newpage
% 
% \doublespacing
% % \onehalfspacing
% 
% 
% % \thispagestyle{empty}

@@

The males-to-females ratio at birth increased dramatically over the last three 
decades in India as access to prenatal sex determination expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}
Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask about use of 
sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have relied on a simple method for establishing 
factors that affect sex selection: use the sex of children born as the 
dependent variable, and estimate the effects of variables on the probability of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
% Poor nutritional and health status of women may lead to 
% relatively more girls being born, probably as an evolutionary
% response to maximise reproductive success \citep{trivers73,wells00}.
% Empirical evidence from Ethiopia and from cross-country data supports
% this hypothesis \citep{klasen02,gibson03}.
% % , although acute under-nutrition may not have this effect \citep{stein04}.
% See also the discussion for India in \citet{bhaskar07}.
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

@@

\section{Theoretical Framework\label{sec:model}}

[TK Calling this a model will be overkill - save the model for the fertility paper]

@@

The model I present here serves two purposes.
First, the model informs the empirical strategy presented below.
Second, I show that there are cases where parents reconsider their decision to 
use sex selection between births, which underscores why 
using only observed births may provide a biased view of how much
sex selection takes place and what the final sex ratio will be.

[move discussion of spacing from footnote above to here?]

This is obviously a simplification and since the interaction
between son preference and spacing is at the core of this paper,
I first discuss factors that may affect spacing between birth.%
\footnote{
It is possible that parents care directly about the
length of spacing between births, but 
spacing outcomes are more likely the results of parents weighing many
factors against each other.
}

One factor that might impact desired spacing is health.
As discussed above, shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.%
\footnote{
\citet{wolpin84} and \citet{Newman1988} present dynamic models 
where both births and deaths are stochastic, but the likelihood of
a child dying is independent of the length of spacing to
the prior birth.
}
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing is too short.

@@

Finally, increased reliability of access and effectiveness of
contraceptive can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of 
contraception---resulting in longer spacing---to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may also help explain shorter spacing for better 
educated women than for less educated women, provided that   
knowledge and ability to use contraception differ
across education groups \citep{Tulasidhar1993,Whitworth2002}.

% "Solving" the model

Instead of trying to model each of the factors that may affect
spacing I instead set up a dynamic model with a discount rate.
Parents maximize utility over a finite time horizon $T$
\begin{equation}
\max E \sum_{t=0}^T d^t [ 
u(c_t) + \alpha u^b(b_t) + (1-\alpha) u^g(g_t) 
],
\label{eq:utility_1}
\end{equation}
where $E$ is the expectation operator,
$c_t$ consumption in period $t$, 
$b_t$ the stock of boys in period $t$,
$g_t$ the stock of girls in period $t$, and
$d$ the discount rate.
For ease of analysis I assume that utility in each period is separable in 
parental consumption, boys, and girls.
I define son preference, $\alpha$, as higher marginal utility of next child 
being a boy if parents have an equal number, possibly zero, of boys and girls.
The utility function covers a variety of different preferences 
from equal utility of boys and girls ($\alpha = 0.5$ provided
$u^b(\cdot) = u^g(\cdot)$) to an extreme 
son preference where there is no utility of having a daughter 
($\alpha = 1$).%
\footnote{
Other ways of defining son preference, such as wanting at least one son 
and not caring about sex composition after that, are possible but 
will be explored here.
The basic results remain as long as there is different
utility of sons and daughters for at least some sex compositions.
}
The lower $d$ is the more parents prefer immediate utility relative
to future utility.
Hence, the discount factor is a convenient way to capture how patient parents 
are and therefore how willing parents are to wait through multiple abortions 
to ensure a son.

At the beginning of each period parents have 3 possible choices:
no pregnancy, a pregnancy without a prenatal sex determination scan,
or a pregnancy with a prenantal sex determination scan followed by an
abortion if the fetus is female.
The probability of conceiving a boy is $\pi$, which is
the same for all parents and in all periods.
There is no child mortality, so both
$b$ and $g$ are non-decreasing over time.

If a child is born parents incur a cost $p_k$ for
that child in the period of birth and all subsequent periods
until the end of the time horizon.%
\footnote{
Clearly boys and girls may carry different costs, but since 
the model already assumes that
parents have different preferences over boys and
girls there is no reason to further complicate the model.
Similarly, the cost of maintaining a child likely depend on 
its age, but incorporating that would add little to 
the conclusions.
}
If parents select a pregnancy with prenatal sex determination
they pay $p_s$ and also $p_k$ if the fetus is a son
and therefore carried to term.
For two reasons the prenatal sex determination scan is best thought of as
a bundled service that includes both the prenatal sex determination scan 
\emph{and} the abortion if the fetus is female.
First, it is more tractable than including additional steps in
the decision process, especially since the only motivation for
a prenatal scan in the model is sex selection, and a scan would
therefore be ``wasted'' unless a female fetus was aborted.
Second, I argue in the online Appendix that both parents and providers 
may prefer a bundled service, where there is an upfront cost and the service 
includes both the prenatal screening and the abortion---if needed---from the 
same provider.
The budget constraint for each period is then
\begin{equation}
I_t = c_t + p_k(b_t + g_t) + p_s s,
\end{equation}
where $I_t$ is income in period $t$
and
$s$ is an indicator variable for whether the 
parents purchased prenatal sex determination.%
\footnote{
This assumes that there is no saving or borrowing
across periods.
This is a common assumption in this type of model.
See, for example, 
\citet{Heckman1976},
\citet{wolpin84},
\citet{rosenzweig85},
\citet{Newman1988},
and
\citet{Leung1991}.
That there are only two stock variables, the number
of boys and the number of girls, greatly simplifies the analysis.
}

This maximization problem can be solved using dynamic programming, 
but because the choice variable and the state variables (number of
boys and girls) are discrete a closed-form solution
is generally not available.
% There are two discrete state variables (boys and girls) and one
% choice variable with three potential choices (no pregnancy, pregnancy without prenatal sex 
% determination, and pregnancy with prenatal sex determination).
% Combined with the finite time horizon a problem of this type is especially suited for
% simulation and the model is simulated using backward induction.
This means that simulation is the best way to analyze the model,
but before turning to simulations it is worthwhile illustrating
how the model works.
I begin with the possible utility outcomes in period $T$, the last period 
in which parents make decisions on fertility.
If parents decide on ``no pregnancy'' their utility in the final period will be
\begin{equation}
U_T^N =
u(I - p_k(b_{T-1} + g_{T-1})) + \alpha u^b(b_{T-1}) + (1 - \alpha) u^g(g_{T-1}).
\end{equation}
If they instead decide to have a pregnancy without a prenatal scan 
their expected utility will be 
\begin{multline}
U_T^B = 
\pi [ 
u(I - p_k(b_{T-1} + g_{T-1} + 1)) + \alpha u^b(b_{T-1}+1) + (1-\alpha) u^g(g_{T-1}) 
]
+\\
(1-\pi) [ 
u(I - p_k(b_{T-1} + g_{T-1} + 1)) + \alpha u^b(b_{T-1}) + (1-\alpha) u^g(g_{T-1}+1) 
].
\end{multline}
Finally, if they decide to use prenatal sex determination their
expected utility will be
\begin{multline}
U_T^S = 
\pi [ 
u(I - p_k(b_{T-1} + g_{T-1} + 1) - p_s) + \alpha u^b(b_{T-1}+1) + (1-\alpha) u^g(g_{T-1}) 
]
+ \\
(1-\pi) [ 
u(I - p_k(b_{T-1} + g_{T-1}) - p_s) + \alpha u^b(b_{T-1}) + (1-\alpha) u^g(g_{T-1}) 
],
\end{multline}
since parents incur the cost of the scan, $p_s$, whether they conceive
a male or a female fetus, but will not incur any additional childrearing costs
if a girl is conceived.

To illustrate how the decision process works, consider
the decision in the final period on whether to use sex selection or not
(assuming that both are superior to no birth).
Parents choose a pregnancy without prenatal sex determination
if $U_T^B > U_T^S$, which leads to
\begin{multline} \label{eq:last_simple}
(1-\pi)(1-\alpha) [ u^g(g_{T-1}+1) - u^g(g_{T-1}) ]
> \\
\pi u(I - p_k(b_{T-1} + g_{T-1} + 1) - p_s) 
+ (1-\pi) u(I - p_k(b_{T-1} + g_{T-1}) - p_s) \\
- u(I - p_k(b_{T-1} + g_{T-1}+1)).
\end{multline}
% This can also be written as
% \begin{multline} \label{eq:last_full}
% (1-\pi)(1-\alpha) [ u^g(g_{T-1}+1) - u^g(g_{T-1}) ]
% > \\
% \pi [u(I - p_k(b_{T-1} + g_{T-1} + 1) - p_s) - u(I - p_k(b_{T-1} + g_{T-1}+1))] \\
% + (1-\pi) [ u(I - p_k(b_{T-1} + g_{T-1}) - p_s) - u(I - p_k(b_{T-1} + g_{T-1}+1))].
% \end{multline}
Hence, parents choose a pregnancy without prenatal scan over
a pregnancy with a prenatal scan in the final period if
the additional utility of having an extra girl (the left-hand-side),
weighted by the probability of having a girl and the weight parents
assign to utility from having girls, is larger than the expected impact on
utility from consumption (the right-hand side), where
the first two terms are the expected utility from consumption if 
using prenatal sex determination and the last term is utility from
consumption if parents simply have a birth.%
\footnote{
Notice that the marginal utility of having a boy does not enter
into the decision on whether to use prenatal sex determination or not
(except through the value of $\alpha$).
Only the marginal utility of having another girl
and the total number of children, which affects the trade-off between
consumption and use of sex selection, matters.
This is an artifact of the separability between consumption, number of
boys, and number of girls in the utility function.
}
% With some abuse of language I will call this difference the 
% marginal utility of a child, in this case a girl.

@@

Finally, increasing the cost of children, $p_k$---as would happen 
with higher education of mothers---increases the likelihood that sex 
selection will be used at a given number of children.%
\footnote{
Higher education likely also increases household income, which
makes it \emph{less} likely parents will choose a birth with prenantal 
scan over a birth without a prenatal scan.
Higher income means lower marginal utility from consumption, which makes the 
whole right-hand side move toward zero.
For a given number of girls, this makes it more likely 
that the utility from an additional girl is sufficiently high that parents 
will simply have a birth.
Empirically, increasing education of mothers is, however, strongly associated 
with lower fertility, which indicates that 
the negative effect of higher opportunity cost more than outweighing 
the positive effect of higher income \citep{schultz97}.
}
There are two reasons for this.
First,  when the cost per child goes up the given number of children 
translates into less income per period available for consumption.
Second, the decline in consumption from having an additional
child becomes larger.
These combine to increase the marginal utility of consumption and thereby 
make it more attractive to have a prenatal scan.

% % Theory on whether reversal of use
% A question of interest is when parents would pick
% a scan in one period and no scan in the following period.
% For this to happen parents must choose scan, conceive a girl, and then
% have a birth without a scan.
% The available choices in the second to last period are the same as in the last
% period: no birth, pregnancy without a scan, and a pregnancy with a scan.
% Let $V_T(B)$ be the expected utility of the optimal choice in the last
% period if the outcome of the choice in period $T-1$ is a boy.
% Similarly, $V_T(G)$ and $V_T(N)$ are the expected utilities from
% having a girl and no birth in period $T-1$.
% In period $T-1$ parents will choose to have a pregnancy without a scan
% over a pregnancy with a scan if
% \begin{multline}
% (1-\pi)(1-\alpha) [ u^g(g_{T-2}+1) - u^g(g_{T-2}) ] 
% > \\
% \pi u(I - p_k(b_{T-2} + g_{T-2} + 1) - p_s) \\
% + (1-\pi) [ u(I - p_k(b_{T-2} + g_{T-2}) - p_s) + d( V_T(N)- V_T(G))] \\
% - u(I - p_k(b_{T-2} + g_{T-2} + 1)).
% \end{multline}
% As before the left-hand side is the marginal utility of
% an additional daughter, but the right-hand side now also 
% incorporates the discounted value of the difference between expected utility 
% in the last period if no children are born in $T-1$ and expected utility in 
% the last period if a girl is born in $T-1$.
% 
% For parents to change their decision on abortion it must be
% the case that the right-hand side for $T-1$ is larger than 
% the left-hand side \emph{and} the left-hand side need to 
% be larger than the right-hand side without the discounted value
% of the difference in expected utility.
% Note that the left-hand side remains the same because with
% a scan no girls are born and the parents therefore enter the
% last period with the same number of girls as they exited
% period $T-2$.%
% \footnote{
% This would not be the case if utility was not separable between
% boys and girls.
% }
% This, in turn, requires that $V_T(N) - V_T(G)$ is positive.
% Identifying when this holds becomes complicated and the
% level of complication increases as we move towards to the
% beginning of the time horizon.

% [Simulation of model]
Although it is possible to examine how decisions are made in the last period,
the absence of closed-form solutions means that I cannot directly 
examine important questions such as whether parents will reverse their
use of sex selection between births.
I therefore now turn to simulation of the model for different parameters.
Appendix Tables \ref{tab:simul_fertility_100}-\ref{tab:simul_fertility_85} 
show the simulation results for average  fertility and the ratio of boys to births for 
selected values of son preference ($\alpha$), cost of prenatal sex determination ($p_s$) 
and cost of children ($k$).
Appendix Tables \ref{tab:simul_abortion_100}-\ref{tab:simul_abortion_85}
show the corresponding simulation results for number of abortions per 100 women, 
average number of spells where prenatal scan has been used, 
and the percent of spells where the choice is initially
pregnancy \emph{with} a prenatal scan but the birth is from a pregnancy \emph{without}
a prenatal scan.
The latter captures the likelihood of changing the decision to use sex selection
between births and I cover that separately below.
Appendix Tables \ref{tab:simul_spacing_r100_a05}-\ref{tab:simul_spacing_r100_a065}
show the average and maximum spacing between births by parity.

The specification and parameter values are chosen to illustrate
the effects of changing parameters, rather than to provide
a realistic set of outcomes.
The utility function used for the simulations is%
\footnote{
See \citet{leung94} for a discussion of potential utility
functions with different preferences for sons and daughters.
}
\[
U = 0.27\frac{(c-30)^{0.5}}{0.5} + \alpha \frac{b^{0.75}}{0.75} + (1-\alpha) \frac{g^{0.75}}{0.75}.
\]
The simulations presented in the tables assume 15 periods, a period income of 
$I = 200$, and that the probability of a son is 0.5.
The four parameters and their examined values are the cost of 
children ($k \in 12, 14, 16, 18, 20, 22$), 
cost of prenatal sex determination ($p \in 12, 16, 20, 24,  200$)---where 
the last value is equivalent to no prenatal sex determination available,
the degree of son preference ($\alpha \in 0.5, 0.55, 0.6, 0.65$),
and the discount rate ($d \in 0.85, 0,9, 0.95, 1.0$).
The optimal choice sets are applied to a sample of 50,000 ``individuals'', each 
with 15 ``potential'' children, whose sex comes from a binomial distribution 
with the probability above.%
\footnote{
I also ran simulations on an expanded set of parameter values.
I use these results for regression analysis of the effect of parameters.
These are available upon request.
}

% Access / cost of prenatal sex determination

Lowering the cost of prenatal sex determination does, however, bias the overall sex 
ratio further towards boys, and the stronger the son preference and the 
more patient the parents, the larger the change in sex ratio.
Correspondingly, both the number of abortions per woman and the
average number of spells per woman where a prenatal scan is used also increases.
Since fertility changes little, the increased use of abortion
is associated with longer spacing between births.
This increase in spacing is, however, not uniform across parities.
In addition to increasing the overall sex ratio, lower cost of prenatal 
scan leads to use of sex selection at earlier parities.
The result is that in some cases there is either no change or a \emph{decrease} 
in average spacing for a given parity---although with a more than compensating 
increase in an earlier parity.%
\footnote{
One example is the second parity for $k= 22$ and $\alpha = 0.65$
when cost of prenatal scan decreases from 24 to 20.
The average spacing for the second parity falls from 1.5 to 1.25,
but the average spacing for the first parity goes from 1 (no
use of sex selection) to 1.5.
Another example is for $\alpha = 0.5$ when cost of prenatal scan falls 
from 20 to 16 when cost of children is 16. 
The average spacing for parity 5 births falls from 1.44 to 1.18, but
the spacing for parity 4 births increases only from 1 to 1.09. 
What this hides is, of course, that there are substantially more 
parity 4 births than there are parity 5 births--and only a slight 
decline in fertility from 4.16 to 4.11---and that the number of
abortions per 100 woman doubles. 
}
Hence, it is possible for individual parities to show a \emph{decline}
in sex selection at the same time as the overall use goes up.
This is an important caution against over-interpreting
individual parity results.
It also highlights the need to incorporate fertility progression into
estimates, something the simple model is incapable of.

@@

An important question here is whether parents ever
change their use of prenatal sex determination between
two consecutive births.
I capture this by the percent of spells where prenatal
scans were initially used for one or more period, but the birth
is the result of a pregnancy without a prenatal scan.
Whether parents will consider the duration from last birth too long 
depends mainly on the discount rate.
On average, the more impatient parents are---the less they
like long spells between births---the more likely they are to
change their decision to use sex selection.%
\footnote{
Using only those situations where there are any abortions,
regressing the discount rate against the percent spells where
prenatal scan is initially used but not for the birth show
a strong positive effect---higher number, more patient---holding
all the other parameters constant.
}
Lowering the cost of prenatal sex determination generally leads to 
a smaller percentage of parents who change their use of sex selection.
But, the total number of spells with a change may still
go up because lower cost of prenatal sex determination also
increases the use of sex selection.%
\footnote{
An example of this is when the discount rate is 0.95, cost of
children 14, cost preference of 0.6, and the cost of a prenatal scan falls 
from 16 to 12, resulting in a decline in the probability of changing 
from 38.1\% to 30.9\%.
In this case the average number of spells with a change increases
from 0.19 spells at the higher cost of prenatal scan 
to 0.37 for the lower cost.
}

@@

Increasing the cost of children and the degree of son preference
both tend to lower the likelihood that parents will change their
decision to use sex selection.
Working in the opposite direction is that some women with higher cost
of children, for example because of higher education, are also likely 
to marry later.
The later marriage effectively shortens their reproductive time horizon
and thereby makes it more likely that they will reverse their use of 
sex selection between births.
% Again, the caveat is that extensive use of sex selection
% can lead parents close to the end of the time horizon, at which
% point they may change their decision.

% Hence, this shows that parents may reverse the use of sex selection, 
% both when they use sex selection relatively infrequently and when they 
% use it heavily.

% Furthermore, the reversal of use of sex selection is 
% often followed by a spell where sex selection is used again,
% especially in the case when a girl is born.

% What are the model highlights 
There are three main conclusions from the model.%
\footnote{
A caveat is that these results reflect the particular utility 
function chosen.
If parents have a different utility function, like a preference
for one son but otherwise no preference across boys and girls
the results may look different.
% As shown it still is possible, however, to get sex selection
% for balance and for parents to change their decision on use
% of sex selection between births.
}
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
% Furthermore, as the simulations show it is possible for son 
% preference to decline while the sex ratios becoming more unequal 
% provided the cost of prenatal scan decreases and/or cost of children 
% increases sufficiently (or people become more patient).
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

@@

The theoretical model has three implications for fertility decisions with access to 
sex-selective abortions: 
a higher probability that next child is a son, 
longer average time to next birth, 
and that the use of prenatal sex determination may change between one observed birth and 
the next.
All three must be incorporated into the estimation strategy
to provide unbiased and precise estimates of the use of sex-selective abortions.%
\footnote{
For comparison the online Appendix presents the estimation strategy 
for the simple model.
The online Appendix also shows how individual parity results compare 
between the simple model and the model suggested here.
As mentioned, the simple model cannot provide complete fertility 
and total number of abortions predictions.
}

The empirical model is a discrete time non-proportional competing risk hazard model with 
two exit states: either a boy or a girl is born.%
\footnote{
\cite{merli00} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
A woman's reproductive life is divided into spells which each covers the period between
births (from marriage to first birth for the first spell).
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$ when either a birth occurs or the spell is censored.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, if the data are transformed so the unit of analysis is spell unit rather 
than the individual woman, the model can be estimated using a standard multinomial logit model.%
\footnote{
% [problem with IIA assumption (independence of irrelevant alternatives]
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
% Even though boys and girls are not perfect substitutes they are 
% very similar and this may present a problem for the IIA assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
% In other words, the assumption requires that the hazard of having a boy 
% relative to not having a child is uncorrelated with the corresponding
% hazard of having a girl relative to not having a child.
% An important unobservable factor is fecundity.
% Assume that a couple has low fecundity.
% Even with a strong son preference, low fecundity  makes the couple less 
% likely to engage in sex-selective abortions because they will have
% difficulty conceiving again and having a girl might be preferable to having
% no children.
% This is exacerbated by the potential for a further lowering of fecundity
% as a direct result of using abortions.
% Fecundity therefore affect both the hazards of having children and the
% relative likelihood of having a boy versus a girl.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
% Finally, what is of interest here is the relationship between the covariates 
% and the rate of occurrence of the different exit types rather than the interrelation between the failure
% types or what would happen if one of the exit types were removed
% making the independence assumption less problematic.%
% \footnote{
% As shown by \citet{kalbfleisch02} data with spell length, exit states and
% fixed covariates does not allow one to distinguish between an independent 
% competing risk model and an infinitude of dependent models that lead to the
% same type-specific hazards.
% }
}
In the reorganized data the outcome variable is zero if the
woman does not have a child in a given period, one if she gives birth to a son in that period,
and two if she gives birth to a daughter in that period.

@@

% \begin{figure}[htp]%
% \centering
% \includegraphics[width=0.4\textwidth]{sex_ratio_marriage}
% \caption{Ratio of boys in ``first'' births
% \label{fig:sex_ratio_recall}}
% \end{figure}

@@

% [NEW OUTLINE]
% - Desired fertility analysis
% - changes over time for well-educated women (spell 2 and 3)
% - comparison across groups (no, some, high education, for spells 1, 2, and 3)

% - 1st spell for all individuals (compare with other research)
% - High education group for spell 2 and above
% - Medium education group for spell 2 and above
% - Low education group for spell 2 and above
% - Predicted changes in fertility and number of abortion and comparison with standard model

@@

% % [Seems to be a repeat of what is in estimation strategy]
% The survival curves show the estimated probability of not having had a birth
% yet by length of the spell.
% The survival curves are included for two reasons.
% First, they show the progression to the next birth by quarter and how quickly it occurs,
% if at all.
% Second, they provide ``weighting'' for the associated percentage boys born.
% The steeper the survival curve, the more weight a given spell period has because it is 
% based on more births, whereas a flat survival curve means little weight because the 
% percentage boys is based on few births.

@@

% [CLOSING ARGUMENTS ON COMPARISON?]
% [logit will tend to underestimate when sex-selective abortions is spreading/increasing
% and overestimate when sex-selective abortions is declining.]

>>>>

% sex-selective Abortion
% Begun originally: 01/11/02
% Begun.: 2017-04-03
% Edited: 2017-04-03

@@

\setcounter{page}{-1}
\maketitle
\thispagestyle{empty}

% \setcounter{page}{0}
\newpage

\doublespacing
% \onehalfspacing

% \thispagestyle{empty}

@@

The males-to-females ratio at birth increased dramatically over the last three 
decades in India as access to prenatal sex determination expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth \citep{Yi1993,park95}.
}
Our understanding of the use of sex selection is, however, constrained by 
lack of information;
there are no official data on sex selection, and the few surveys that ask about use of 
sex selection show signs of serious under-reporting \citep{goodkind96}.
With no direct information, researchers have relied on a simple method for establishing 
factors that affect sex selection: use the sex of children born as the 
dependent variable, and estimate the effects of variables on the probability of having a son.%
\footnote{
In the absence of any interventions, the probability of having a son
is approximately 0.512, which is independent of genetic factors \citep{ben-porath76b,jacobsen99}.
With fetus sex random, a statistically significant variable indicates an association with 
use of sex-selective abortions.
Examples of studies that have used this approach are \cite{retherford03b},
\cite{jha06}, and \cite{abrevaya09}. 
}
Based on this simple method, we know that families with no sons are more likely to
use sex selection the higher the parity;
that use of sex selection increases with socioeconomic status, especially education;
and that sex selection is more widespread in cities than in rural areas 
\citep{retherford03b,jha06,abrevaya09}.%
\footnote{
There is substantial disagreement on whether sex-selective abortion is used for the 
first birth \citep{retherford03b,jha06}.
}

@@

The method can better predict ``final'' sex ratios than the simple model

\section{The Interaction between Son Preference and Birth Spacing\label{sec:model}}

The extension of birth spacing that result from abortions as described above is 
only one part of the interaction between son preference and birth spacing.

- the decision on sex selection and spacing
    - the mechanical extension
    - the possibility of changing use of sex selection

- what other factors might change birth spacing?
    - increased use of contraceptives
    - changes in labor force participation

\citep{Soest2012} for a recent example of spacing and son preference

@@

I show that there are cases where parents reconsider their decision to 
use sex selection between births, which underscores why 
using only observed births may provide a biased view of how much
sex selection takes place and what the final sex ratio will be.

As discussed above, shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.%
\footnote{
\citet{wolpin84} and \citet{Newman1988} present dynamic models 
where both births and deaths are stochastic, but the likelihood of
a child dying is independent of the length of spacing to
the prior birth.
}
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing is too short.

@@

[TK this needs to move to later]
Finally, increased reliability of access and effectiveness of
contraceptive can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of 
contraception---resulting in longer spacing---to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may also help explain shorter spacing for better 
educated women than for less educated women, provided that   
knowledge and ability to use contraception differ
across education groups \citep{Tulasidhar1993,Whitworth2002}.

To illustrate how the decision process works, consider
the decision in the final period on whether to use sex selection or not
(assuming that both are superior to no birth).
Parents choose a pregnancy without prenatal sex determination
if $U_T^B > U_T^S$, which leads to
\begin{multline} \label{eq:last_simple}
(1-\pi)(1-\alpha) [ u^g(g_{T-1}+1) - u^g(g_{T-1}) ]
> \\
\pi u(I - p_k(b_{T-1} + g_{T-1} + 1) - p_s) 
+ (1-\pi) u(I - p_k(b_{T-1} + g_{T-1}) - p_s) \\
- u(I - p_k(b_{T-1} + g_{T-1}+1)).
\end{multline}
Hence, parents choose a pregnancy without prenatal scan over
a pregnancy with a prenatal scan in the final period if
the additional utility of having an extra girl (the left-hand-side),
weighted by the probability of having a girl and the weight parents
assign to utility from having girls, is larger than the expected impact on
utility from consumption (the right-hand side), where
the first two terms are the expected utility from consumption if 
using prenatal sex determination and the last term is utility from
consumption if parents simply have a birth.%
\footnote{
Notice that the marginal utility of having a boy does not enter
into the decision on whether to use prenatal sex determination or not
(except through the value of $\alpha$).
Only the marginal utility of having another girl
and the total number of children, which affects the trade-off between
consumption and use of sex selection, matters.
This is an artifact of the separability between consumption, number of
boys, and number of girls in the utility function.
}

@@

Lowering the cost of prenatal sex determination does, however, bias the overall sex 
ratio further towards boys, and the stronger the son preference and the 
more patient the parents, the larger the change in sex ratio.
Correspondingly, both the number of abortions per woman and the
average number of spells per woman where a prenatal scan is used also increases.
Since fertility changes little, the increased use of abortion
is associated with longer spacing between births.
This increase in spacing is, however, not uniform across parities.
In addition to increasing the overall sex ratio, lower cost of prenatal 
scan leads to use of sex selection at earlier parities.
The result is that in some cases there is either no change or a \emph{decrease} 
in average spacing for a given parity---although with a more than compensating 
increase in an earlier parity.
Hence, it is possible for individual parities to show a \emph{decline}
in sex selection at the same time as the overall use goes up.
This is an important caution against over-interpreting
individual parity results.
It also highlights the need to incorporate fertility progression into
estimates, something the simple model is incapable of.

@@

An important question here is whether parents ever
change their use of prenatal sex determination between
two consecutive births.
I capture this by the percent of spells where prenatal
scans were initially used for one or more period, but the birth
is the result of a pregnancy without a prenatal scan.
Whether parents will consider the duration from last birth too long 
depends mainly on the discount rate.
On average, the more impatient parents are---the less they
like long spells between births---the more likely they are to
change their decision to use sex selection.%
\footnote{
Using only those situations where there are any abortions,
regressing the discount rate against the percent spells where
prenatal scan is initially used but not for the birth show
a strong positive effect---higher number, more patient---holding
all the other parameters constant.
}
Lowering the cost of prenatal sex determination generally leads to 
a smaller percentage of parents who change their use of sex selection.
But, the total number of spells with a change may still
go up because lower cost of prenatal sex determination also
increases the use of sex selection.

@@

% What are the model highlights 
There are three main conclusions from the model.%
\footnote{
A caveat is that these results reflect the particular utility 
function chosen.
If parents have a different utility function, like a preference
for one son but otherwise no preference across boys and girls
the results may look different.
}
The increased use of sex selection shows up both in the 
proportion of boys among births and longer spacing between 
births---and the higher the parity the longer the duration between 
births.
The other factors, such as the price of sex selection, 
discount factor, and son preference also affect the use
of sex selection, but to a much smaller degree.
Second, parents do reverse the use of sex selection, 
both when they use sex selection relatively infrequently and when 
they use it heavily.
Reversal are more likely the more impatient parents are or the
shorter time they have left in their reproductive horizon, which
means that the longer the duration has been since the last birth
the more likely they are to reverse their decision to use sex selection.
Finally, the model shows that son preference in the number of
children is not required for an unequal sex ratio.
As long as parents are only willing to abort female fetuses, declining
fertility will lead to increased use of sex selection to ensure 
a balanced number of sons and daughters.

@@

The theoretical model has three implications for fertility decisions with access to 
sex-selective abortions: 
a higher probability that next child is a son, 
longer average time to next birth, 
and that the use of prenatal sex determination may change between one observed 
birth and the next.
All three must be incorporated into the estimation strategy
to provide unbiased and precise estimates of the use of sex-selective abortions.%

The empirical model is a discrete time non-proportional competing risk hazard model with 
two exit states: either a boy or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
A woman's reproductive life is divided into spells which each covers the period between
births (from marriage to first birth for the first spell).
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$ when either a birth occurs or the spell is censored.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, if the data are transformed so the unit of analysis is spell unit rather 
than the individual woman, the model can be estimated using a standard multinomial logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is zero if the
woman does not have a child in a given period, one if she gives birth to a son in that period,
and two if she gives birth to a daughter in that period.

@@

@@

@@

@@
